One-dimensional Classical Gas in a Harmonic Trap with Repulsive Interaction

### 2017-12-05 draft_05.tex
almost done with breathing mode section

\begin{document}
 \title{One-dimensional Classical Gas in a Harmonic Trap with Repulsive Interaction}
 \author{Zhiyu}
 \affiliation{Fudan University, Shanghai, China}

@@

Our model for the interacting classical gas involves particles with a simple finite-range repulsive
interaction.  Two particles repel each other whenever they are within a distance $\sigma$ from each
other.  The force of repulsion is a constant, $F_0$, within this distance and zero when the distance
is larger. 
%
\begin{equation}
F(|x|) ~=~ \begin{cases}
F_0  &\quad |x|<\sigma\\
0  &\quad |x|>\sigma
\end{cases}
\label{eq:preparation1}
\end{equation}
where $|x|$ is the distance between two particles.  This equation describes the magnitude of the
interparticle force; the direction is always repulsive.

The gas contains $N$ such identical particles, each of mass $m$, in a harmonic trap.  The
Hamiltonian describing the gas is 
\begin{equation}
H=\frac{1}{2}m\omega_0^2\sum_{i}x_i^2+\frac{1}{2}m\sum_i
v_i^2+\sum_{\left|x_i-x_j\right|\textless\sigma}F_0\left(\sigma-\left|x_i-x_j\right|\right)
\label{eq:preparation2}
\end{equation}
%
It is useful to rescale the quantities.  We will measure distance, time, energy and force in units
of $\sigma$, $1/\omega$, $m\omega_0^2\sigma^2$ and $m\omega_0^2\sigma$ respectively:
\begin{equation} \label{eq:transform}
  \tilde{x_i} = \frac{x_i}{\sigma}\ , \quad 
  \tilde{H} = \frac{H}{m\omega_0^2\sigma^2}\ ,  \quad \tilde{t} = \omega_0t \ ,  \quad 
\tilde{F_0} = \frac{F_0}{m\omega_0^2\sigma}\ .
\end{equation}
%
Eq.~\ref{eq:preparation2} is then rewritten as
\begin{equation}
\tilde{H}=\frac{1}{2}\sum_{i}\tilde{x_i}^2+\frac{1}{2}\sum_i\left(\frac{d\tilde{x}_i}{d\tilde{t}}\right)^2+\sum_{\left|\tilde{x_i}-\tilde{x_j}\right|\textless 1}F_0\left(1-\left|\tilde{x_i}-\tilde{x_j}\right|\right)
\end{equation}
%
Through this rescaling, we have reduced the number of the parameters in our model to three: the
energy $\tilde{H}$, the interaction strength $\tilde{F_0}$, and the number of particles $N$.  The
rescaling is equivalent to setting $\sigma$, $\omega_0$ and $m$ to $1$; this is what we do in our
numerical simulation.  In the rest of this paper, we will use ``E" and ``$F_0$" to denote the
reduced versions $\tilde{H}$ and $\tilde{F_0}$.

We will be concerned with the size of the cloud, which we quantify through the root-mean-square of
particle positions  $\left\lbrace x_i\right\rbrace$
\begin{equation}
R\equiv\left(\overline{x_i^2}\right)^\frac{1}{2}
\label{eq:def_of_R}
\end{equation}
and refer to as the `radius' or `cloud radius'.

In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units, $E\approx NR^2$.

\begin{figure}[tbhp]
\centering
\includegraphics[scale=0.6]{ZhiyuPictures/N=5_GS_pre_2_rev.eps}
\caption{The ground state of a system with $N=5$ particles, shown via the positions of each
  particle, labeled 1 through 5.  At large interactions, the particles position themselves just
  outside the range of interactions of the neighboring particles.}
\label{fig:GS1}
\end{figure}

@@

At smaller $F_0$, the energy is minimized by having the particles approximately equidistant, at
distance $\approx 2F_0/\omega_0^2$ from each other.  In this `solid'-like state, the low-lying
excitation involves independent oscillation of the particles around their equilibrium position.

In Fig.\ref{fig:GS1}, we can see this crossover from the ``solid-like" limit (left) to the
``hardcore gas" limit (right).

@@

Figures \ref{fig:Breathingfrequency2_0} and \ref{fig:Breathingfrequency2_1} show such phase space
snapshots.  We generally start with particles distributed at random positions around the trap
center, initially with no velocity (zero momentum).  The initial state thus has the points lined up
along the $X$ axis.  In the absence of interactions, each particle would undergo simple harmonic
oscillation with period $T=2\pi/\omega_0 = 2\pi$.  The point corresponding to each particle than
executes clockwise elliptical motion around in the $X$-$P$ plane.  (In our units, the elliptical
trajectory is actually circular.)  This results in the initial line remaining a straight line and
rotating clockwise with exact period $T$.  The effect of interaction is to smear out the line and
spread the points out toward a rotationally invariant distribution in the  $X$-$P$ plane.  The top
panel of Figure  \ref{fig:Breathingfrequency2_0} shows this during the first period and Figure
\ref{fig:Breathingfrequency2_1} shows this process over a much longer timescale.

@@

In the absence of interactions, each point (each particle) is stationary in the rotating-frame
phase-space picture.  As seen in lower panel of Figure \ref{fig:Breathingfrequency2_0} and in Figure
\ref{fig:Breathingfrequency2_1}, interactions cause a gradual distortion of the line as well as some
degree of rotation.
%
The rotation that is visible in the already rotating $X_{\rm rot}$-$P_{\rm rot}$ frame is the
interaction-induced shift of the breathing frequency from the noninteracting value,
$\omega=2\omega_0$.  The distortion of the line toward an eventually rotationally invariant
distribution may be regarded as thermalization or ergodicity.  In the next sections we explore these
two interaction-induced effects.

@@

Since the number of particles we use in the Molecular Dynamics simulation is limited (5 to 50), the
random noise is significant. If we start with a state whose configuration in phase-space is an
ellipse slightly deviated from circle, the oscillating amplitude will be too small to be
distinguished from noise. Therefore, we start from an extreme case -- line distribution (random x,
zero p) and observe the oscillation of R(t).

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h]
\center
\includegraphics[scale=0.32]{ZhiyuPictures/freq_scanF_differentE_log_2_with_oscillation_demo.png}
\caption{Breathing mode frequency measured at different $E$ and $F_0$, for $N=5$ particles. Inset: a
  demonstration of the oscillation of $E_{trap}$, the total potential energy of particles in the
  trap, Since $E_{trap}\propto \sum_{is}{x_i^2}$, it manifests the oscillation behavior of the
  radius $R(t)$ as defined in Eq.~\ref{eq:def_of_R}.  The breathing frequency is obtained from the
  Fourier transform of $R(t)$.}
\label{fig:Breathingfrequency1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

We are interested in the oscillations of the size $R(t)$ of the cloud.  It is convenient that our
finite-size simulations start with a line distribution in phase space.  If we started with a state
whose configuration in phase-space is only slightly different from a circle, the oscillating
amplitude would be too small to be distinguished from noise.

Without interaction, the breathing mode frequency is exactly 2.  This is visualized readily from the
top panel of \ref{fig:Breathingfrequency2_0}, where $R(t)$ is the extent of the distribution in the
horizontal ($X$) direction.  As the line rotates clockwise with frequency $\omega=1$, within each
period the line is twice horizontally aligned (maximum $R$) and twice vertically aligned (minimum
$R$), so that the frequency of $R(t)$ is $\omega_B=2$.  When there is interaction, the frequency
will get shifted: $\omega_B = 2+\delta$.  We measure the radius of the cloud $R(t)$ and get the
frequency spectrum of its oscillation behavior by Fourier transform. Then we take the peak frequency
near 2 as the breathing mode frequency.

@@

Below, we will provide a detailed phase-space argument for these behaviors.  However, a simple
real-space picture explains both effects qualitatively as well.  For very large $F_0$, when the
interaction is `hard-core'-like, two particles exchange momentum immediately at a collision.  By
exchanging the labels of the particles during the collision, this can be interpreted as follows:
particle $A$ carrying momentum $P_A$ jumps by distance $\sigma$ to the right, while particle $B$
carries its momentum $P_B$ and jumps by distance $\sigma$ to the left.  In this manner, every
collision will save a particle some time, $\frac{\sigma}{v}$.  This translates into an increase of
the breathing frequency.  The speed $v$ per particle is on average $\sim\sqrt{E/N}$, and the number
of collision each particle experiences in one period of harmonic oscillation is $\sim N$; hence we
obtain $\delta\sim N^{\frac{3}{2}}E^{-\frac{1}{2}}$.  When $F_0$ is smaller, we can no longer think
of the particles as hard rods; the collisions now take finite time during which the speeds of the
particles are actually slowed down and then sped up again, as they either cross paths or bounce from
each other.  When $F_0$ is small enough, this approaching time will at some critical value of $F_0$
consume the time saved by the finite range of the interaction.  This explains why there is a
small-$F_0$ regime for which the shift $\delta$ is negative.

\subsection{Phase space estimate}

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h
]
\subfigure[]{
\begin{minipage}[b]{0.4\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/bounce.png}
\end{minipage}
}
\subfigure[]{
\begin{minipage}[b]{0.4\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/pass.png}
\end{minipage}
}
\caption{a schematic diagram of the collision process: \textbf{(a)}When $F_0\sigma$ is large, particles can not pass each other. Subduing to a constant repulsive force in collision, their trajectory in rotating frame is a circular arc.  \textbf{(b)} When $F_0\sigma$ is small, particles easily pass each other, so that the force they feel is not continuous. Their trajectory has a turning point. The dashed bright blue trajectory is identical with the one in (a), which is the critical situation between passing and bouncing. In that case, particles have zero relative velocity when they collide. }
\label{fig:Breathingfrequency3}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

To analyze the rotation relative to the $X_{\rm rot}$-$P_{\rm rot}$ frame, we consider two-particle
collisions.  In Figure \ref{fig:Breathingfrequency3}(a), we show schematics of such a collision near
the trap center at strong interactions.  For very large $F_0$, the particles simply change momentum
when they are at distance $\sigma$ from each other; this process is shown in red as two straight
lines.  (The particles are initially at $X=\pm\frac{1}{2}\sigma$ and remain at these $X$ values,
but exchange their momenta.)   For smaller $F_0$, there is change of both position and momentum, as shown in yellow, green
and blue for successively weaker interactions. The blue line shows a case where $F_0$ is smaller
enough such that the particles can actually cross each other.

In Figure \ref{fig:Breathingfrequency3}(a) we show collisions for small $F_0$, so that the particles
pass each other.  in the $X_{\rm rot}$-$P_{\rm rot}$ plane, this is seen as a turning point in the trajectory.

We now estimate $\delta$ at large $F_0$.  The relevant collision process is that shown by the red
lines in Figure \ref{fig:Breathingfrequency3}(a).  The two particles, which are initially on the
$X_{\rm rot}$ axis, are moved to two other points, on the thick gray line, which deviates a small
angle $\alpha$ away from original configuration.  In every period of the harmonic oscillation, each
particle meets each of the other particles twice.  Half of these collisions (about $N$ collision)
are between particles with large difference in momentum, which is the process shown in
Fig.\ref{fig:Breathingfrequency3}.  (As for the remaining half of the collisions, colliding
particles have smaller difference in their momentume, e.g., both particles having positive
momentum.  In this estimate we  ignore the effects of these collisions, as they clearly contribute
far less to the rotation of the particle distribution.)

The precession angle is $\alpha\sim \sigma/P$.  In our rescaled units, the momenta that the
particles have near $x=0$ are of the order $R$, the size of the cloud.  (In phase space $R$ can be
interpreted either as the extent in real space when the velocities are small, or as the extent in
momentum space when the particles approach $x=0$, i.e. when the distribution is along the $P$
direction.)  Thus, we can estimate  $\alpha\sim \sigma/R$, i.e.,   $\alpha\sim 1/R$, since our unit
of length is $\sigma$.  This is the rotation per unit collision.  Since there are $\sim N$
collistions per period $T=2\pi$, we have the estimate for the interaction-induced rotation per unit
time as
\begin{equation}
\delta ~=~ 2 \frac{N}{2\pi} \frac{1}{R} ~=~  \frac{1}{\pi}  N^{3/2}E^{-1/2}
\label{eq:breathingfrequency1}
\end{equation}
The factor $2$ accounts for the fact that each rotation of the elongated cloud in phase space
corresponds to two breathing mode periods.  In the last step we have used the estimate $E\sim
NR^2$.

@@

%% \subsection{Estimating Breathing Frequency}

%% Let us begin with two-particle case. When interaction is very strong, the effect of interaction is
%% equivalent to exchanging momentum when two particles approach each other at distance $\sigma$. In
%% the rotating frame in phase-space, this process can be interpret as follows:

%% (See Fig.\ref{fig:Breathingfrequency3}(a))Two particles flip to another line(thick gray line) which
%% deviate a small angle $\alpha$ away from original configuration. Similar process happen when there
%% are more particles. In every period of harmonic oscillation, each particle meet N particles and
%% collide 2N times, while half of the collisions (N times) are between particles with huge difference
%% in momentum, which is just the process shown in Fig.\ref{fig:Breathingfrequency3}. (As for the
%% remaining half of collisions, colliding particles have small difference in their momentum. Let us
%% say, both particles have positive momentum. In Fig.\ref{fig:Breathingfrequency3}, it can be shown
%% that two particles on the same half of the P-axis seldom contribute to the rotation of
%% distribution.)

%% Let us estimate the precession angle $\alpha$ in one collision event by $\frac{\sigma}{R}$, which
%% follows from the two-particle case study. Radius of the cloud R can be estimated according to
%% $E=Nm\omega_0^2R^2$ We will get the precession angular velocity
%% \begin{equation}
%% \delta=2\omega_0\frac{N^\frac{3}{2}m^{\frac{1}{2}}E^{-\frac{1}{2}}\sigma\omega_0}{2\pi}
%% \label{eq:breathingfrequency1}
%% \end{equation}

When $F_0$ is not that large, the frequency behavior seems complicated (see Fig.\ref{fig:Breathingfrequency1}). Usually, when $F_0$ increase, frequency goes down below 2 first and then rise up and converge to some value higher than 2. The reason could also be well understood with the mechanism described before. In the discussion above, we go to high $F_0$ limit, which means particles exchange momentum in infinitesimal time. For $F_0$ is not very big, the finite interaction time should be taken into consideration. During this time, the motion of each particle in phase-space is exactly moving along the direction of the real P axis at 'velocity' $F0/m$. Since the real P axis is rotating counter-clockwise, the particle will follow a circular trajectory(orange and green dashed line in Fig.\ref{fig:Breathingfrequency3}) the $\alpha$ that we once use to estimate the precession motion should decrease with the increase of interaction time. There should be some ``critical point" where $\alpha$ goes from plus to minus (dashed green line). It is easy to see that the contribution of each collision event to the phase-space precession of the cloud is negative in Fig.\ref{fig:Breathingfrequency3})(b). So at this point, we will see the precession  slows down or even be reversed, which means the breathing frequency is 2 or a little bit lower than 2.

\begin{figure}[h]
\centering
\includegraphics[scale=0.6]{ZhiyuPictures/freq_scanF_scanE_pre_Font18.eps}
\caption{dependence of breathing frequency on Energy, N=20. Different color stands for different $F_0$: Red for $F_0=2\times10^3$; Yellow for $F_0=3\times10^3$; Green for $F_0=1\times10^4$;  Blue for $F_0=1\times10^5$. Run with three different initial states in each case, and in each run, measure the frequency in 0-100 time unit(label as crosses) and in 900-1000 time unit (label as circles). The black dashed line is the prediction of Eq.\ref{eq:breathingfrequency1}.}
\label{fig:Breathingfrequency4}
\end{figure}

The dependence of frequency on energy and $F_0$ is shown in Fig.\ref{fig:Breathingfrequency4}. For the first thing, as expected when $F_0$ is high the frequency behavior is well-described by linear prediction Eq.\ref{eq:breathingfrequency1}. Secondly, the figure shows the frequency's dependence on $F_0$. As we explained, when $F_0$ is not so large, the precession angle decrease. In language of real space, the finite $F_0$ create a finite approaching time, thus consumes the time saved by finite-range collision. As a result, the frequency should be lower when $F_0$ decrease, and reach 2 earlier.

the critical point that frequency decreases to 2 can be estimate by:
\begin{equation}
\omega_0 \tau=\alpha
\end{equation}
which means that the approaching time $\tau$ (estimated by $R/F_0$) exactly cancels the precession angle we derived before. One can show that this condition is equivalent to
\begin{equation}
\frac{E}{N}=F_0\sigma
\end{equation}
which can also be understood as the criterion of whether two particles will bounce or pass by if one estimate the average internal energy of a pair with $\frac{E}{N}$.

>>>>

\begin{document}
\title{One-dimensional Classical Gas in a Harmonic Trap with Repulsive Interaction}

\author{
Zhiyu Dong,$^{1,2}$
R. Moessner,$^2$
and Masudul Haque$^{2,3}$}

\affiliation{
  $^1$ Fudan University, Shanghai, China\\ 
  $^2$ Max Planck Institute for the Physics of Complex Systems, N{\"o}thnitzer Str.~38, 01187 Dresden, Germany\\
  $^3$ Department of Theoretical Physics, Maynooth University, Co. Kildare, Ireland
}

@@

Our model for the interacting classical gas involves particles with a simple finite-range repulsive
interaction.  Two particles repel each other whenever they are within a distance $\sigma$ from each
other.  The force of repulsion is a constant, $F_0$, within this distance and zero when the distance
is larger. 
%
\begin{equation}
\Big| F(|x|) \Big| ~=~ \begin{cases}
F_0  &\quad |x|<\sigma\\
0  &\quad |x|>\sigma
\end{cases}
\label{eq:preparation1}
\end{equation}
where $|x|$ is the distance between two particles.  This equation describes the magnitude of the
interparticle force; the direction is always repulsive.

The gas contains $N$ such identical particles, each of mass $m$, in a harmonic trap.  The
Hamiltonian describing the gas is 
\begin{equation}
H=\frac{1}{2}m\omega_0^2\sum_{i}x_i^2+\frac{1}{2}m\sum_i
v_i^2+\sum_{\left|x_i-x_j\right|\textless\sigma}F_0\left(\sigma-\left|x_i-x_j\right|\right)
\label{eq:preparation2}
\end{equation}
%
It is useful to rescale the quantities.  We will measure distance, time, energy and force in units
of $\sigma$, $1/\omega$, $m\omega_0^2\sigma^2$ and $m\omega_0^2\sigma$ respectively:
\begin{equation} \label{eq:transform}
  \tilde{x_i} = \frac{x_i}{\sigma}\ , \quad 
  \tilde{H} = \frac{H}{m\omega_0^2\sigma^2}\ ,  \quad \tilde{t} = \omega_0t \ ,  \quad 
\tilde{F_0} = \frac{F_0}{m\omega_0^2\sigma}\ .
\end{equation}
%
Eq.~\ref{eq:preparation2} is then rewritten as
\begin{equation}
  \tilde{H} =
  \frac{1}{2}\sum_{i}\tilde{x_i}^2
  + \frac{1}{2}\sum_i\left(\frac{d\tilde{x}_i}{d\tilde{t}}\right)^2 
  + \sum_{\left|\tilde{x_i}-\tilde{x_j}\right|\textless
    1}\tilde{F}_0\left(1-\left|\tilde{x_i}-\tilde{x_j}\right|\right)  
\end{equation}
%
Through this rescaling, we have reduced the number of the parameters in our model to three: the
energy $\tilde{H}$, the interaction strength $\tilde{F_0}$, and the number of particles $N$.  The
rescaling is equivalent to setting $\sigma$, $\omega_0$ and $m$ to $1$; this is what we do in our
numerical simulation.  In the rest of this paper, we will use ``E" and ``$F_0$" to denote the
reduced versions $\tilde{H}$ and $\tilde{F_0}$.

We will be concerned with the size of the cloud.  We quantify the size through the root-mean-square
of particle positions $\left\lbrace x_i\right\rbrace$, 
\begin{equation}
R\equiv\left(\overline{x_i^2}\right)^\frac{1}{2} \, ,
\label{eq:def_of_R}
\end{equation}
and refer to this quantity as the `radius' or `cloud radius'.

In the dilute gas limit, there are few or no interactions occuring at most times.  Thus the energy
$E$ is dominated by the trap energy, i.e., $E\approx m\omega_0^2\sum_{i}x_i^2= Nm\omega_0^2R^2$, or,
in our units,
\begin{equation}
  E\approx NR^2 \, .
  \label{eq:E_and_R}
\end{equation}
This is a good approximation at large $F_0$ because, for very large interactions, the particles
behave has hard rods which bounce off each other very rapidly, so that at most instants one can
expect no collisions to be taking place.  It is also expected to be a good approximation at very
small $F_0$ since we can then simply neglect the interaction energy.  Thus, we will use this as a common
estimate for $R$ in terms of the energy.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbhp]
\centering
\includegraphics[scale=0.6]{ZhiyuPictures/N=5_GS_pre_2_rev.eps}
\caption{The ground state of a system with $N=5$ particles, shown via the positions of each
  particle, labeled 1 through 5.  At large interactions, the particles position themselves just
  outside the range of interactions of the neighboring particles.}
\label{fig:GS1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

At very small $F_0$, the energy is minimized by having the particles approximately equidistant, at
distance $\approx 2F_0/\omega_0^2$ from each other.  In this `solid'-like state, the low-lying
excitation involves independent oscillation of the particles around their equilibrium position.

In Fig.\ref{fig:GS1}, we can see this crossover from the ``solid-like" limit (left) to the
``hardcore gas" limit (right).  In between, there is a rich staircase-like structure, as the
particles attempt to minimize the interaction by being at distance $>1$ from as many other particles
as allowed by the trap energy.

@@

Figures \ref{fig:Breathingfrequency2_0} and \ref{fig:Breathingfrequency2_1} show such phase space
snapshots.  We generally start with particles distributed at random positions around the trap
center, initially with no velocity (zero momentum).  The initial state thus has the points lined up
along the $X$ axis.  In the absence of interactions, each particle would undergo simple harmonic
oscillation with period $T=2\pi/\omega_0 = 2\pi$.  The point corresponding to each particle executes
clockwise elliptical motion around the $X$-$P$ plane.  (In our units, the elliptical trajectory is
actually circular.)  This results in the initial line remaining a straight line and rotating
clockwise with exact period $T$.  The effect of the interaction is to smear out the line and spread
the points out toward a rotationally invariant distribution in the $X$-$P$ plane.  The top panel of
Figure \ref{fig:Breathingfrequency2_0} shows this during the first period and Figure
\ref{fig:Breathingfrequency2_1} shows this process over a much longer timescale.

@@

In the absence of interactions, each point (each particle) is stationary in the rotating-frame
phase-space picture.  As seen in lower panel of Figure \ref{fig:Breathingfrequency2_0} and in Figure
\ref{fig:Breathingfrequency2_1}, interactions cause a gradual distortion of the line as well as some
degree of rotation.
%
The rotation that is visible in the already rotating $X_{\rm rot}$-$P_{\rm rot}$ frame is the
interaction-induced shift of the breathing frequency from the noninteracting value,
$\omega_B=2\omega_0$.  The distortion of the line toward an eventually rotationally invariant
distribution may be regarded as thermalization or ergodicity.  In the next sections we explore these
two interaction-induced effects.

@@

We use the Verlet algorithm (molecular dynamics) to numerically simulate the cloud, using particle
numbers between 5 and 50.  Our force is simple, so that calculating the force at each step is
inexpensive, however, the theta function dependence of the force on particle positions requires the
use of fine timesteps at the beginning and end of each collision process.

The initial state is taken to have particles with zero velocity and random positions; hence the line
distribution in the phase space picture.  This has the advantage that the breathing motion is
prominently visible.  In addition, the question of long-time relaxation has the simple
interpretation of evolving from the line distribution to a circularly symmetric distribution.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbp]
\center
\includegraphics[scale=0.32]{ZhiyuPictures/freq_scanF_differentE_log_2_with_oscillation_demo.png}
\caption{Breathing mode frequency measured at different $E$ and $F_0$, for $N=5$ particles. Inset: a
  demonstration of the oscillation of $E_{trap}$, the total potential energy of particles in the
  trap, Since $E_{trap}\propto \sum_{is}{x_i^2}$, it manifests the oscillation behavior of the
  radius $R(t)$ as defined in Eq.~\ref{eq:def_of_R}.  The breathing frequency is obtained from the
  Fourier transform of $R(t)$.}
\label{fig:Breathingfrequency1}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

We are interested in the oscillations of the size $R(t)$ of the cloud.  It is convenient that our
finite-size simulations start with a line distribution in phase space.  If we started with a state
whose configuration in phase-space deviates only slightly from circular symmetry, the oscillating
amplitude would be too small to be distinguished from noise.

Without interaction, the breathing mode frequency is exactly 2.  This is visualized readily from the
top panel of \ref{fig:Breathingfrequency2_0}, where $R(t)$ is the extent of the distribution in the
horizontal ($X$) direction.  As the line rotates clockwise with frequency $\omega_0=1$, within each
period the line is twice horizontally aligned (maximum $R$) and twice vertically aligned (minimum
$R$), so that the frequency of $R(t)$ is $\omega_B=2$.  When there is interaction, the frequency
will get shifted: $\omega_B = 2+\delta$.  We measure the radius of the cloud $R(t)$ and get the
frequency spectrum of its oscillation behavior by Fourier transform. Then we take the peak frequency
near 2 as the breathing mode frequency.

@@

Below, we will provide a detailed phase-space argument for these behaviors.  However, a simple
real-space picture explains both effects qualitatively as well.  For very large $F_0$, when the
interaction is `hard-core'-like, two particles exchange momentum immediately at a collision.  By
exchanging the labels of the particles during the collision, this can be interpreted as follows:
particle $A$ carrying momentum $P_A$ jumps by distance $\sigma$ to the right, while particle $B$
carries its momentum $P_B$ and jumps by distance $\sigma$ to the left.  In this manner, every
collision will save a particle some time, $\frac{\sigma}{v}$.  This translates into an increase of
the breathing frequency.  The speed $v$ per particle is on average $\sim\sqrt{E/N}$, and the number
of collision each particle experiences in one period of harmonic oscillation is $\sim N$; hence we
obtain $\delta\sim N^{\frac{3}{2}}E^{-\frac{1}{2}}$.  When $F_0$ is smaller, we can no longer think
of the particles as hard rods; the collisions now take finite time during which the speeds of the
particles are actually slowed down and then sped up again, as they either cross paths or bounce from
each other.  When $F_0$ is small enough, this approaching time will at some critical value of $F_0$
consume the time saved by the finite range of the interaction.  This explains why there is a
small-$F_0$ regime for which the shift $\delta$ is negative.  Figure \ref{fig:Breathingfrequency1}
shows that this critical value of $F_0$, where $\delta$ changes sign and $\omega_B$ crosses $2$,
increases with energy.

\subsection{Estimates using the rotating phase space}

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h
]
\subfigure[]{
\begin{minipage}[b]{0.4\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/bounce.png}
\end{minipage}
}
\subfigure[]{
\begin{minipage}[b]{0.4\textwidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/pass.png}
\end{minipage}
}
\caption{Schematic diagrams of two-particle collision process in the $X_{\rm rot}$-$P_{\rm rot}$
  plane. \textbf{(a)}When $F_0$ is large, particles cannot pass each other.  Being subject to a
  constant repulsive force, their trajectories during collision are circular arcs.  For very strong
  interactions (red dashed line, straight), they exchange momenta instantaneously.  The yellow,
  green, and blue dashed curves are for successively weaker interactions.  \textbf{(b)} When $F_0$
  is small, particles pass each other, so that the force they feel is not continuous.  Their
  trajectory has a turning point. The dashed bright blue trajectory is identical with the one in
  (a), which is the critical situation between passing and bouncing. In that case, particles have
  zero relative velocity when they collide. }
\label{fig:Breathingfrequency3}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

To analyze the rotation relative to the $X_{\rm rot}$-$P_{\rm rot}$ frame, we consider two-particle
collisions.  In Figure \ref{fig:Breathingfrequency3}(a), we show schematics of such a collision near
the trap center at strong interactions.  For very large $F_0$, the particles simply exchange momentum
when they are at distance $\sigma$ from each other; this process is shown in red as two straight
lines.  (The particles are initially at $X=\pm\frac{1}{2}\sigma$ and remain at these $X$ values,
but exchange their momenta.)   For smaller $F_0$, there is change of both position and momentum, as shown in yellow, green
and blue for successively weaker interactions. The blue line shows a case where $F_0$ is smaller
enough such that the particles can actually cross each other.

In Figure \ref{fig:Breathingfrequency3}(a) we show collisions for small $F_0$, so that the particles
pass each other.  The force changes direction discontinuously when the particles cross.  In the
$X_{\rm rot}$-$P_{\rm rot}$ plane, this is seen as a sharp turning point in the trajectory.

We now estimate $\delta$ at large $F_0$.  The relevant collision process is that shown by the red
lines in Figure \ref{fig:Breathingfrequency3}(a).  The two particles, which are initially on the
$X_{\rm rot}$ axis, are moved to two other points, on the thick gray line, which deviates a small
angle $\alpha$ away from original configuration.  In every period of the harmonic oscillation, each
particle meets each of the other particles twice.  Half of these collisions (about $N$ collisions)
are between particles with large difference in momentum, which is the process shown in
Fig.\ref{fig:Breathingfrequency3}.  (As for the remaining half of the collisions, colliding
particles have smaller difference in their momenta.  In this estimate we ignore the effects of these
collisions, as they clearly contribute far less to the rotation of the particle distribution.)

The precession angle is $\alpha\sim \sigma/P$.  In our rescaled units, the typical momenta of
interacting partilces is of the order $R$.  (In phase space $R$ can be interpreted either as the
extent in real space when the velocities are small, or as the extent in momentum space when the
particles approach $x=0$, i.e. when the distribution is along the $P$ direction.)  Thus, we can
estimate $\alpha\sim \sigma/R$, i.e., $\alpha\sim 1/R$, since our unit of length is $\sigma$.  This
is the rotation per unit collision.  Since there are $\sim N$ collistions per period $T=2\pi$, we
have the estimate for the interaction-induced rotation per unit time as
\begin{equation}
\delta ~=~ 2 \frac{N}{2\pi} \frac{1}{R} ~=~  \frac{1}{\pi}  N^{3/2}E^{-1/2}
\label{eq:breathingfrequency1}
\end{equation}
The factor $2$ accounts for the fact that each rotation of the elongated cloud in phase space
corresponds to two breathing mode periods.  In the last step we have used the estimate $E\sim
NR^2$.

The effect of interaction on breathing frequency is more complicated at smaller $F_0$.  As we have
seen in Figure \ref{fig:Breathingfrequency1}, the effect of a small $F_0$ is to \emph{decrease} the
breathing frequency from its non-interacting value $\omega_B=2$.  This can also be understood using
phase-space pictures of collision processes.  For $F_0$ is not very big, the finite interaction time
needs to be taken into consideration.  During this time, the motion of each particle in phase-space
is moving exactly along the direction of the real $P$ axis at the rate $F_0/m$. Since the real $P$
axis is rotating counter-clockwise, the particle will follow a circular trajectory, e.g., following
the orange or green dashed lines in Fig.~\ref{fig:Breathingfrequency3}(a).  The angle $\alpha$ that
we used above to estimate $\delta$ thus decreases with the increase of interaction time; it is
smaller for the yellow line and even opposite for the green line.  This explains the negative
contribution of interaction to $\delta$, for small enough values of $F_0$.  At even smaller values
of $F_0$, shown in Fig.~\ref{fig:Breathingfrequency3})(b), the particles cross each other.  The
resulting final values are such that the line joining the post-interaction locations of the two
particles has negative $\alpha$, i.e., is tilted clockwise with respect to the $X_{\rm rot}$ axis,
meaning a negative contribution to the breathing frequency.

We can also estimate the critical value of $F_0$ for a given energy (or alternatively the critical
value of $E$ for a given $F_0$) between positive and negative contributions to the breathing
frequency.  Since the $P$ axis rotates with frequency $\omega_0=1$ in the $X_{\rm rot}$-$P_{\rm
  rot}$ plane, the post-collistion position of the particle in the $X_{\rm rot}$-$P_{\rm rot}$ plane
makes angle $\omega_0\tau$ with the red line in Figure \ref{fig:Breathingfrequency3}(a).  Here
$\tau$ is the time over which the interaction acts.  The crossover between positive and negative
$\delta$ is found by comparing this angle to $\alpha$:
\begin{equation}
\omega_0 \tau=\alpha
\end{equation}
The time $\tau$ is approximately the time it takes for the momentum to change sign due to a constant
force $F_0$, since the momentum is $\sim R$, this means $\tau\sim R/F_0$.  Using our previous
estimate  $\alpha\sim \sigma/R$, together with $E\sim NR^2$, gives us the condition
\begin{equation}
  \frac{E}{N} \sim F_0
  \label{eq_freq_turning_condition}
\end{equation}
for the breathing frequency to cross the non-interacting value $\omega_B=2$.  This is roughly the
same criterion for whether two particles will bounce or cross each other in a typical collision.

The condition \eqref{eq_freq_turning_condition} is consistent with Figure
\ref{fig:Breathingfrequency1}, where we noted that the critical $F_0$ grows with increasing energy.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h]
\centering
\includegraphics[scale=0.6]{ZhiyuPictures/freq_scanF_scanE_pre_Font18.eps}
\caption{dependence of breathing frequency on Energy, N=20. Different color stands for different $F_0$: Red for $F_0=2\times10^3$; Yellow for $F_0=3\times10^3$; Green for $F_0=1\times10^4$;  Blue for $F_0=1\times10^5$. Run with three different initial states in each case, and in each run, measure the frequency in 0-100 time unit(label as crosses) and in 900-1000 time unit (label as circles). The black dashed line is the prediction of Eq.\ref{eq:breathingfrequency1}.}
\label{fig:Breathingfrequency4}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

In Figure \ref{fig:Breathingfrequency4} we show a quantitative test of the main rotating-phase-space
prediction, Eq.~\eqref{eq:breathingfrequency1}.  For large $F_0$, the frequency indeed scales as
$E^{-1/2}$ and for $F_0=10^5$ is quite close to the actual numerical values prediction
\eqref{eq:breathingfrequency1}.  Even for smaller $F_0$, at small enough energies (larger $E^{-1/2}$
values) the breathing mode becomes proportional to  $E^{-1/2}$.

@@
